---
title: "Project2"
Date: created on feb 21st, 2017
output: html_document
Author: Ken Gu
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
```

Studies on the effect of weather events on personal as well as property damages in US based on NOAA database.

1. Synopsis: 
  1.1 Background: Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

  1.2 DataSource Of Analysis: 
This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

  1.3 Summary:
The report leverages line plot to indicate the 2 following questions:
    1.3.1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
    1.3.2. Across the United States, which types of events have the greatest economic consequences?

2. Data Processing:
This section describes the how the data is loaded.
RawData -> Pre-processed data -> Plot
Step 1: read rawData
Step 2: clean the rawData; filter out defined EVTYPE from the document; only select columns: EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP; 
Step 3: re-organize the step2 data to calculate the total harmful events to population health; economic consequences


```{r, cache=TRUE}
stormRawData <- read.csv("./StormData.csv", header=TRUE)
processData <- select(stormRawData, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP )
# clean up EVTTYPE
evtTypeList <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
processData <- filter(processData, EVTYPE %in% toupper(evtTypeList))
processData[,1] <- sapply(processData[,1],function(row) iconv(row, "latin1", "ASCII", sub=""))

# filter out missing Values
compData1 <- complete.cases(processData[,1], processData[,2], processData[,3])
compData2 <- complete.cases(processData[,1], processData[,4], processData[,5], processData[,6], processData[,7])

# clean up EVTTYPE
processData1 <- processData[compData1,]
processData2 <- processData[compData2,]
grpResult1 <- group_by(processData1, EVTYPE)

# summarize the results for health costs
result1 <- summarise(grpResult1, sum(as.numeric(FATALITIES)) + sum(as.numeric(INJURIES)))
colnames(result1) <- c("EVTYPE", "SUMNUM")

# make the order
result1 <- result1[order(-result1$SUMNUM),]
result1 <- slice(result1, 1:5)
result1


```
3. Results:
```{r, cache=TRUE}
# draw graphs for question 1
ggplot(result1, aes(EVTYPE, SUMNUM)) + geom_bar(stat = "identity", fill = "blue")
maxname1 <- filter(result1, SUMNUM==max(result1[,2]))[,1]

# for question 2
processData2 <- mutate(processData2, ecoCost=0)

processData21 <- filter(processData2, toupper(processData2$PROPDMGEXP)=="K")
processData22 <- filter(processData2, toupper(processData2$PROPDMGEXP)=="M")
processData23 <- filter(processData2, toupper(processData2$PROPDMGEXP)=="B")
processData24 <- filter(processData2, toupper(processData2$CROPDMGEXP)=="K")
processData25 <- filter(processData2, toupper(processData2$CROPDMGEXP)=="M")
processData26 <- filter(processData2, toupper(processData2$CROPDMGEXP)=="B")
processData21[,8] <- processData21[,4]*1000
processData22[,8] <- processData22[,4]*1000000
processData23[,8] <- processData23[,4]*1000000000
processData24[,8] <- processData24[,6]*1000
processData25[,8] <- processData25[,6]*1000000
processData26[,8] <- processData26[,6]*1000000000

processDataC2 <- rbind(processData21, processData22, processData23, processData24,  processData25,processData26)
grpResult2 <- group_by(processDataC2, EVTYPE)
result2 <- summarise(grpResult2, sum(ecoCost) )
colnames(result2) <- c("EVTYPE", "SUMECOCOST")
#reorder result2
result2 <- result2[order(-result2$SUMECOCOST),]
result2 <- slice(result2, 1:5)
result2
ggplot(result2, aes(EVTYPE, SUMECOCOST)) + geom_bar(stat = "identity", fill = "blue")

```



4. Conclusion: 
    
    4.1 Across the United States, event - TORNADO - is most harmful with respect to population health.
    
    4.2 Across the United States, event - FLOOD -  have the greatest economic consequences


